Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-06-24 paper/main.tex
associativity

\subsection{Equivariant Transformations}
Along with preserving topology, visualization algorithms are expected to preserve the structure of the data fields, i.e. the column types in a table, when they translate data to visual representations. The mathematical term for this is equivariance and an equivariant map is a map that transforms input with structure to output with compatible structure. Bertin codified the idea that visual encoding of a data field should match the structure of the field; for example ordered data should be encoded as an ordered visual element or quantitative data as a quantitative visual element \cite{bertinSemiologyGraphicsDiagrams2011a}. This structure mapping is necessary to ensure that the visual representations in graphs are not distorting the data\cite{tufteVisualDisplayQuantitative2001} and graphs are easier to understand when properties match \cite{norman_things_smart}. The expectation of matching structure is generalized as a mapping of a binary operator that acts on the data domain to a binary operator that acts on the graphic domain by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}. An invertible binary operator is the basis of Kindlemannn and Scheideggers's algebraic model of visualization\cite{kindlmannAlgebraicProcessVisualization2014}. They propose three specific types of equivariance: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, that changing the data should correspond to changes in the visualization in a perceptually significant manner. In \autoref{sec:atct:sheaves} we introduce a uniform abstract data representation layer, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi}, and the definition of equivariance introduced in \autoref{sec:artist:equivariance} includes the correspondence between data and visual changes. In this work, we formally specify the conditions necessary to build structure preserving components using category theory. A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

@@

\subsubsection{Topology: Base Space \dbase}
We encode the underlying connectivity of the data as the \textcolor{base}{base space} \dbase\ of the fiber bundle because it provides a uniform language for describing the topology regardless of how the data is subset. The base space \dbase\ is a topological space $(\dbase,\mathcal{T}_{\dbase})$, which means it consists of the set of points ${\dbasepoint \in dbasec}$ and topology $\mathcal{T}_{\dbase}$. A topology $\mathcal{T}_{\dbase}$ is a collection of open sets $\openset$ such that the empty set and \dbase are in $\mathcal{T}_{\dbase}$,and every union and intersection of opensets in $\mathcal{T}_{\dbase}$ are also in $\mathcal{T}_{\dbase}$ \cite{bradleyTopologyCategoricalApproach2020}. An openset \openset\ is a generalization of an open interval to a region, meaning an open set contains points inside a bounded region on \dbase and excludes the points on the boundary \cite{TopologyVsTopology}. A property of the base space \dbase\ is that it is a quotient topology, ,meaning it has the finest topology of $\dbase$, which  means that it contains the largest possible collection of distinct open sets on $\dbase$ \cite{QuotientSpaceTopology2020}. For our purposes, $\dbase$ is a set of points that encodes the underlying connectivity of the our dataset, for example, $\dbase$ could be the discrete points, interval, or plane in \autoref{fig:related-work:continuity:ktypes}. The topology on $\dbase$ encodes the space of smallest possible subsets of the data, and the union and intersection rules express that these subsets must be combinable into larger subsets of the data.

We encapsulate the notion of a programmatic topology type as the base space category $\mathcal{\dbasec}$. The base space category $\mathcal{\dbasec}$ has open set objects $\opensetc$. The morphisms of the category are inclusions $\iota: \opensetc_1 \rightarrow \opensetc_2$ such that $\opensetc_1 \subseteq \opensetc_2$. By definition of a catagory\cite{barrCategoryTheoryComputing}, morphisms are commutative $(\iota_{1,2}: \openset_{1} \rightarrow \openset_{2}) \circ (\iota_{2,3}: \openset_{2} \rightarrow \openset_{3}) = \iota_{1,3}: \openset_{1} \rightarrow \openset_{3}$. Another property of morphisms is that they are associative such that $\iota_{3,4}\circ(\iota_{2,3}\circ\iota_{1,2}) = (\iota_{3,4} \circ \iota{2,3})\circ \iota_{1,2}$ where $\iota_{3,4}:\openset_{3}\rightarrow \openset_{4}$. Each object also has an identity morphism $1_{\openset_i}: \openset_i \rightarrow \openset_i$. We propose that these categorical types can be mapped to 
programmatic types that describe the data topology and that the inclusion morphisms can be used to assert that the continuity has been preserved when defining subsets of data.

\subsection{Field Types: Fiber Space \dfiber}
We encode the field types of a dataset as the  \textcolor{fiber}{fiber space} \dfiberc of the fiber bundle because it allows us to encode fields of varying types and dimensionalities in a consistent manner. The fiber \dfiber\ is the preimage of the projection function $\pi$ at a point in the base space $\dfiber_{\dbasepoint} = \pi^{-1}(\dbasepoint)$, meaning that all points in a fiber space $\dfiber_{\dbasepoint}$ map to the same point $\dbasepoint$. The fiber space can be used to encode the space of all possible values of a data field and the fibers can be named, as described by Spivak \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}. The space of all possible values is analogous to a CS data type; therefore the fiber space can be used to encode the data types of the data fields. The fiber is also a topological space, meaning there is no restriction on the dimensionality of the field, thus allowing for the encoding of multi-dimensional fields in the same manner as single dimensional fields.

>>>>

\subsection{Data Field Structure}
Along with preserving topology, visualization algorithms are expected to preserve the structure of the data fields, i.e. the column types in a table, when they translate data to visual representations. The mathematical term for this is equivariance and an equivariant map is a map that transforms input with structure to output with compatible structure. Bertin codified the idea that visual encoding of a data field should match the structure of the field; for example ordered data should be encoded as an ordered visual element or quantitative data as a quantitative visual element \cite{bertinSemiologyGraphicsDiagrams2011a}. This structure mapping is necessary to ensure that the visual representations in graphs are not distorting the data\cite{tufteVisualDisplayQuantitative2001} and graphs are easier to understand when properties match \cite{norman_things_smart}. The expectation of matching structure is generalized as a mapping of a binary operator that acts on the data domain to a binary operator that acts on the graphic domain by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}. An invertible binary operator is the basis of Kindlemannn and Scheideggers's algebraic model of visualization\cite{kindlmannAlgebraicProcessVisualization2014}. They propose three specific types of equivariance: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, that changing the data should correspond to changes in the visualization in a perceptually significant manner. In \autoref{sec:atct:sheaves} we introduce a uniform abstract data representation layer, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi}, and the definition of equivariance introduced in \autoref{sec:artist:equivariance} includes the correspondence between data and visual changes. In this work, we formally specify the conditions necessary to build structure preserving components using category theory, because it provides a rich language for expressing the structure of objects, functions between objects, and the constraints that must hold for these functions to compose \cite{wielsManagementEvolvingSpecifications1998,yorgeyMonoidsThemeVariations}. A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

@@

\subsubsection{Topological Structure: Base Space \dbase}
We encode the underlying connectivity of the data as the \textcolor{base}{base space} \dbase\ of the fiber bundle because it provides a uniform language for describing the topology regardless of how the data is subset. The base space \dbase\ is a topological space $(\dbase,\mathcal{T}_{\dbase})$, which means it consists of the set of points ${\dbasepoint \in dbasec}$ and topology $\mathcal{T}_{\dbase}$. A topology $\mathcal{T}_{\dbase}$ is a collection of open sets $\openset$ such that the empty set and \dbase are in $\mathcal{T}_{\dbase}$,and every union and intersection of opensets in $\mathcal{T}_{\dbase}$ are also in $\mathcal{T}_{\dbase}$ \cite{bradleyTopologyCategoricalApproach2020}. An openset \openset\ is a generalization of an open interval to a region, meaning an open set contains points inside a bounded region on \dbase and excludes the points on the boundary \cite{TopologyVsTopology}. A property of the base space \dbase\ is that it is a quotient topology, ,meaning it has the finest topology of $\dbase$, which  means that it contains the largest possible collection of distinct open sets on $\dbase$ \cite{QuotientSpaceTopology2020}. For our purposes, $\dbase$ is a set of points that encodes the underlying connectivity of the our dataset, for example, $\dbase$ could be the discrete points, interval, or plane in \autoref{fig:related-work:continuity:ktypes}. The topology on $\dbase$ encodes the space of smallest possible subsets of the data, and the union and intersection rules express that these subsets must be combinable into larger subsets of the data.

We propose that we can express topological structure programmatically by  constructing data types that encapsulates the class of topological structure-.i.e point, line, plane, network, etc. These data types can be formally specified as objects of a category $\mathcal{\dbasec}$. 
The base space category $\mathcal{\dbasec}$ has open set objects $\opensetc$. The morphisms between objects of the category are inclusions $\opensetc_i \xrightarrow{\iota} \opensetc_j$ such that $\opensetc_i \subseteq \opensetc_j$. Given two morphisms $\opensetc_i \xrightarrow{\iota_{i,j}} \opensetc_j$ and $\opensetc_j \xrightarrow{\iota_{j,k}} \opensetc_j$, these morphisms must compose 
\begin{equation}
  \label{eq:atct:base:cat:comp}
\begin{tikzcd}
  \openset_{i} \arrow[rr, "{\iota_{i,j}}"] \arrow[rrrr, "{\iota_{i,k}}", bend right] &  & \openset_{j} \arrow[rr, "{\iota_{j,k}}"] &  & \openset_{k}
\end{tikzcd}
\end{equation}
such that there exists a function $\opensetc_i \xrightarrow{\iota_{i,k}} \opensetc_k$.  The expected composition of functions is often expressed as commutative diagrams, such as \label{eq:atct:base:cat:comp}. Compositions of morphisms are associative
\begin{align}

 & ((\openset_{i}\xrightarrow{\iota_{i,j}} \openset_{j})) \xrightarrow{\iota_{j,k}} \openset_{k} \xrightarrow{\iota_{k,l}} \openset_{l}\label{eq:atct:cat:as1}
  \\
  = & ((\openset_{j}\xrightarrow{\iota_{j,k}} \openset_{k})\xrightarrow{\iota_{k,l}} \openset_{l})

    \openset_{i}\xrightarrow{\iota_{i,j}} \openset_{j}\label{eq:atct:cat:as2}
\end{align}
\note{this needs to be sorted cause this format ends up w/ the same order}

such that \autoref{eq:atct:cat:as1} and \autoref{eq:atct:cat:as2}  return the same value. In general associativity allows us flexibility in building call stacks of curried functio

 While here we discuss composition, associativity, and identity in the context of $\mathcal{\dbase}$, these are general properties of categories \cite{riehlCategoryTheoryContext,barrCategoryTheoryComputing,fongInvitationAppliedCategory2019}.

\subsection{Data Fields: Fiber Space \dfiber}
We encode the field types of a dataset as the  \textcolor{fiber}{fiber space} \dfiberc of the fiber bundle because it allows us to encode fields of varying types and dimensionalities in a consistent manner. The fiber \dfiber\ is the preimage of the projection function $\pi$ at a point in the base space $\dfiber_{\dbasepoint} = \pi^{-1}(\dbasepoint)$, meaning that all points in a fiber space $\dfiber_{\dbasepoint}$ map to the same point $\dbasepoint$. The fiber space can be used to encode the space of all possible values of a data field and the fibers can be named, as described by Spivak \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}. The space of all possible values is analogous to a CS data type; therefore the fiber space can be used to encode the data types of the data fields. The fiber is also a topological space, meaning there is no restriction on the dimensionality of the field, thus allowing for the encoding of multi-dimensional fields in the same manner as single dimensional fields.
